CUDA-accelerated Genetic Feedforward-ANN Training for Data Mining

### 2010-04-21 doc/report.tex
Add two paragraphs on the degenerate output values phenomenon and how we solved it (standardization of feature values)
git-svn-id: http://svn.scesoc.ca/evolution/trunk@112 fee98b4c-d1c3-4d34-9525-a468159e4a48

The classifier receives as inputs one numeric value for each feature in a data set, all of which are normalized to a value range between 0 and 1.  The classifier outputs a single numeric value, also in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.  These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.  Therefore, the fitness function is not limited to lift at 20\% as presented in this paper.

In our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers. Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any value. Figure~\ref{fig:evolution-lift} demonstrates the chromosome structure and its relation to ANN structure.  This structure is simple yet powerful; individual edges are 'removed' by the genetic algorithm when their weight is set to zero and nodes are 'removed' by the genetic algorithm when their RBF width value is set to zero.  Thus, we achieve the performance gains of a fixed ANN topology (allowing for disciplined memory access) while searching a range of topologies.

@@

>>>>

The classifier receives as inputs real-valued features which are standardized prior to processing by the ANN.  The classifier outputs a single real value, in the range of 0 and 1, which represents the confidence with which the pattern is a member of the positive class.  These confidence values can be used for ranking during lift calculations, or for sensitivity or specificity measures with an arbitrary value chosen as a threshold for positive identification of a pattern.  Therefore, the fitness function is not limited to lift at 20\% as presented in this paper.

In our final design, each candidate ANN is represented by a chromosome which consists of four arrays of real-valued numbers. Edge weights between input nodes and hidden layers, edge weights between hidden and output layers, width values for RBFs at hidden nodes and center values for RBFs at hidden nodes are all encoded, and can take on any real value. Figure~\ref{fig:evolution-lift} demonstrates the chromosome structure and its relation to ANN structure.  This structure is simple yet powerful; individual edges are 'removed' by the genetic algorithm when their weight is set to zero and nodes are 'removed' by the genetic algorithm when their RBF width value is set to zero.  Thus, we achieve the performance gains of a fixed ANN topology (allowing for disciplined memory access) while searching a range of topologies.

@@

Due to the inherent limitations of floating-point (FP) representation of real values, the range of inputs to the ANN plays a crucial part in the numerical stability of the classifier during training and during classification. In particular, our use of exponential functions resulted in an ANN very sensitive to output saturation. In these cases, while the input to the \texttt{expf} function is well within the range of representible FP numbers (take for example feature 11 which has a maximum of 8193), the mathematical output of the exponentiation is too large to be represented in FP. This causes degenerate output values such as infinity or NaN (not a number). These special values propagate through the ANN edges to produce degenerate values at the output of the classifier.

To avoid this issue, we standardize the data before training and before classification. For each feature individually, we estimate the 10\textsuperscript{th} and 90\textsuperscript{th} percentiles of the feature values. Then we shift and scale the values such that 80\% of the resulting values are found between -1 and 1. This standardization method was chosen for its robustness against outliers. After standardization, the range of inputs to \texttt{expf} is narrow enough to always produce a valid FP value at the output. The standardization does not affect the classification itself because there are scaling and shifting parameters within the ANN (input to hidden edge weights and hidden center parameters, respectively) which are already part of the search space.
